% ============================================================
% MASTER FILE — GNNocRoute-DRL Paper
% Journal of Systems Architecture (Elsevier)
% ============================================================
\documentclass[5p,twocolumn]{elsarticle}

\graphicspath{{./figures/}}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{footnote}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage{xcolor}

\journal{Journal of Systems Architecture}

\begin{document}

\begin{frontmatter}

\title{GNNocRoute-DRL: Topology-Aware Adaptive Routing for Network-on-Chip using Graph Neural Networks and Deep Reinforcement Learning}

\author[1]{Le Tien Hieu\corref{cor1}}
\ead{letienhieu@mcs.edu.vn}
\cortext[cor1]{Corresponding author}

\address[1]{Institute of Information Technology, Vietnam National University, Hanoi, Vietnam}

\begin{abstract}
Network-on-Chip (NoC) has become the standard communication infrastructure for modern multi-core processors and AI accelerators. However, conventional deterministic routing algorithms suffer from structural congestion imbalance---our graph-theoretic analysis across seven topologies shows that an $8\times8$ mesh under XY routing achieves a congestion imbalance of 0.475, with central nodes experiencing $2-3\times$ higher betweenness centrality than edge nodes. While adaptive routing algorithms improve performance under non-uniform traffic, they rely on local heuristics that cannot leverage global topology structure.

This paper presents GNNocRoute-DRL, a framework combining Graph Neural Networks (GNN) with Deep Reinforcement Learning (DRL) for topology-aware adaptive routing in NoC. The framework introduces two key components: (1) a GATv2 encoder that learns topology-aware node representations correlating with betweenness centrality ($|r| = 0.978$); and (2) a periodic policy optimization scheme that decouples GNN inference ($\approx 10^6$ cycles) from per-packet routing ($<10$ cycles).

Through BookSim2 cycle-accurate simulations on Mesh $4\times4$ with 100 sample periods and 5 random seeds, we evaluate a precomputed GNN-PPO routing policy against four baseline algorithms. Under uniform traffic, GNNocRoute-PPO achieves comparable latency to XY routing (19.8 vs. 19.6 cycles). Under transpose traffic, the fixed policy exhibits limited generalization (44.0 vs. 20.5 cycles for XY). All algorithms saturate below injection rate 0.1 under hotspot conditions. These findings establish both the promise of GNN-enhanced topology awareness---validated through $|r| = 0.978$ correlation with betweenness centrality---and the challenges of offline policy generalization, motivating future work on fully online DRL agents for NoC routing.
\end{abstract}

\begin{keyword}
Network-on-Chip \sep Graph Neural Networks \sep Deep Reinforcement Learning \sep Adaptive Routing \sep Congestion Management
\end{keyword}

\end{frontmatter}

\input{01-introduction}
\input{02-related-work}
\input{03-method}
\input{04-experiments}
\input{05-results}
\input{06-discussion}
\input{07-conclusion}


\nocite{*}
\bibliographystyle{elsarticle-num}
\bibliography{references}

\end{document}
