% ============================================================
% MASTER FILE — GNNocRoute-DRL Paper
% Journal of Systems Architecture (Elsevier)
% ============================================================
\documentclass[5p,twocolumn]{elsarticle}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{footnote}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage{xcolor}

\journal{Journal of Systems Architecture}

\begin{document}

\begin{frontmatter}

\title{GNNocRoute-DRL: Topology-Aware Adaptive Routing for Network-on-Chip using Graph Neural Networks and Deep Reinforcement Learning}

\author[1]{Le Tien Hieu\corref{cor1}}
\ead{letienhieu@mcs.edu.vn}
\cortext[cor1]{Corresponding author}

\address[1]{Institute of Information Technology, Vietnam National University, Hanoi, Vietnam}

\begin{abstract}
Network-on-Chip (NoC) has become the standard communication infrastructure for modern multi-core processors and AI accelerators. However, conventional deterministic routing algorithms suffer from structural congestion imbalance---our graph-theoretic analysis across seven topologies shows that an $8\times8$ mesh under XY routing achieves a congestion imbalance of 0.475, with central nodes experiencing $2-3\times$ higher betweenness centrality than edge nodes. While adaptive routing algorithms improve performance under non-uniform traffic, they rely on local heuristics that cannot leverage global topology structure.

This paper presents GNNocRoute-DRL, a framework combining Graph Neural Networks (GNN) with Deep Reinforcement Learning (DRL) for topology-aware adaptive routing in NoC. The framework introduces three key components: (1) a GATv2 encoder that learns topology-aware node representations correlating with betweenness centrality ($|r| = 0.978$); (2) a periodic policy optimization scheme that decouples GNN inference ($\approx 10^6$ cycles) from per-packet routing ($<10$ cycles); and (3) a DRL agent architecture for multi-objective routing optimization.

We evaluate the framework through extensive BookSim2 cycle-accurate simulations spanning 252 configurations with latency and power analysis. Results demonstrate that topology-aware adaptive routing reduces hotspot latency by up to 5.4\% compared to XY routing with negligible power overhead. Energy analysis confirms that adaptive routing achieves performance gains through better load distribution rather than additional hardware activity. Cross-topology analysis confirms that GNN embeddings generalize across mesh sizes (Pearson $r \approx 0.96$). These findings establish that GNN-enhanced topology awareness is both feasible and beneficial for NoC routing, providing a foundation for next-generation adaptive routing in many-core architectures.
\end{abstract}

\begin{keyword}
Network-on-Chip \sep Graph Neural Networks \sep Deep Reinforcement Learning \sep Adaptive Routing \sep Congestion Management
\end{keyword}

\end{frontmatter}

\input{01-introduction}
\input{02-related-work}
\input{03-method}
\input{04-experiments}
\input{05-results}
\input{06-discussion}
\input{07-conclusion}

\bibliographystyle{elsarticle-num}
\bibliography{references}

\end{document}
