% ============================================================
\section{Introduction}
\label{sec:intro}
% ============================================================

Network-on-Chip (NoC) has become the dominant communication paradigm for multi-core processors, AI accelerators~\cite{journeys}, and System-on-Chips (SoCs). As the number of integrated cores scales from tens to hundreds~\cite{amd2023,apple2023}, the efficiency of the on-chip interconnection network directly determines overall system performance in terms of latency, throughput, and energy consumption~\cite{dally2004}.

The routing algorithm---which determines the path each packet takes from source to destination---is a critical component of NoC design. Conventional deterministic algorithms such as XY dimension-order routing (DOR)~\cite{dally2004} are widely used due to their simplicity and deadlock freedom. However, deterministic routing always selects the same path for each source-destination pair regardless of network congestion, leading to structural bottlenecks.

Our graph-theoretic analysis across seven topologies reveals the severity of this imbalance. Under XY routing on an $8\times8$ mesh, the congestion imbalance---measured as the coefficient of variation ($\sigma/\mu$) of link utilization---reaches 0.475, indicating that nearly half of all links experience significantly uneven load. Central nodes exhibit betweenness centrality $2-3\times$ higher than edge nodes, creating predictable congestion hot spots.

Adaptive routing addresses this limitation by allowing packets to take alternative paths based on network conditions. Research in adaptive NoC routing has evolved through three generations:

\begin{enumerate}
    \item \textbf{Heuristic adaptive routing}~\cite{dyad2004,ebrahimi2019} uses local congestion thresholds to dynamically select output ports. Heuristic adaptive routing can be further categorized into oblivious routing (e.g., Valiant), which randomizes path selection without congestion feedback; partially adaptive routing (e.g., Planar Adaptive, DyAD), which uses turn-model constraints with local congestion thresholds; and fully adaptive routing (e.g., RCA), which requires virtual channels for complete path flexibility. While simple to implement, these approaches rely on hand-crafted heuristics that cannot capture global traffic patterns.
    
    \item \textbf{Deep Reinforcement Learning (DRL) routing}~\cite{wang2022maar,deepnr2022,garnn2023} employs neural network agents trained via RL to make routing decisions. These methods achieve 20--35\% latency improvement over deterministic routing. However, they use Multi-Layer Perceptron (MLP) encoders that process flattened router states as vectors, discarding the inherent graph structure of the NoC.
    
    \item \textbf{Graph Neural Network (GNN) routing for wide-area networks}~\cite{g-routing2023,drl-gnn2019} applies message-passing neural networks to network routing problems. GNNs naturally operate on graph-structured data and can capture topological relationships. However, these approaches target software-defined networks with millisecond-scale latency requirements, not the nanosecond-scale constraints of on-chip routing.
\end{enumerate}

The fundamental challenge in applying GNN-based routing to NoC is \textbf{inference latency}. A single GNN forward pass on a 64-node graph requires approximately $10^6$ cycles on a CPU, while each NoC routing decision must complete within $5-10$ cycles~\cite{wang2022maar}. This approximately five-order-of-magnitude gap renders per-packet GNN inference infeasible.

For a $4\times4$ mesh, the GNN inference executes asynchronously in the background (requiring approximately $5.5$ ms on a standard CPU). The precomputed weight tables are subsequently updated only in response to structural topology shifts or significant traffic variations. This decoupled approach completely eliminates the neural network inference latency from the critical path of per-packet routing.

\subsection{Contributions}

This paper makes the following major contributions:

\begin{enumerate}
    \item \textbf{Congestion analysis:} A quantitative demonstration across seven NoC topologies revealing that deterministic routing intrinsically creates structurally predictable bottlenecks.

    \item \textbf{GNN-topology correlation:} Strong empirical evidence showing that GATv2 node embeddings naturally correlate with betweenness centrality ($|r| = 0.978$), validating the unsupervised emergence of topology-aware structural representations.

    \item \textbf{Port-scoring architecture:} A novel decoupled inference-execution framework that minimizes the per-packet routing cost to a single-cycle table lookup. This hardware-friendly architecture consumes a mere 0.13\% of the router's power budget utilizing a highly compact 4 KB weight table.

    \item \textbf{Comprehensive evaluation:} An extensive cycle-accurate simulation study encompassing over 700 configurations, 7 baseline algorithms, and 6 distinct traffic patterns. Results demonstrate a 5--16\% reduction in hotspot latency compared to Planar Adaptive routing on fault-free meshes, and a remarkably resilient 2.5\% latency degradation (versus a catastrophic 257\% surge for Planar Adaptive) under a 15\% link failure rate.

    \item \textbf{Fault-tolerance and zero-shot generalization:} The introduction of a fault-aware training regime that sustains near-zero throughput loss under 15\% random link failures. Furthermore, the model exhibits zero-shot transferability from $4\times4$ to $8\times8$ mesh scales, achieving 99.78\% minimal-port routing accuracy, providing a strong initialization for few-shot fine-tuning that can reduce training epochs by an estimated 60--80\%.
\end{enumerate}
