% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

This paper presented GNNocRoute-DRL, a efficient, topology-aware adaptive routing framework for Network-on-Chip architectures. By uniquely decoupling complex Graph Neural Network (GNN) inference from real-time execution, we successfully bridged the gap between deep learning expressiveness and the nanosecond-scale latency constraints of physical NoC routers. Our contributions span four major areas:

\textbf{First}, we demonstrated through graph-theoretic analysis across seven topologies that deterministic routing creates structurally predictable congestion bottlenecks. \textbf{Second}, we provided empirical evidence that GATv2-based encoders intrinsically learn topology-aware node embeddings that strongly correlate with betweenness centrality ($|r| = 0.978$), capturing structural dynamics that standard heuristics ignore. \textbf{Third}, we introduced a novel 4-directional port-scoring architecture that reduces per-packet routing to a single-cycle table lookup. This hardware-friendly approach consumes only 0.13\% of the router's power budget via a compact 4 KB weight table. \textbf{Fourth}, we validated the framework's robustness and scalability through a rigorous fault-aware training pipeline and zero-shot transfer evaluations.

Extensive cycle-accurate BookSim2 evaluations demonstrate that GNNocRoute outperforms state-of-the-art baselines. Under hotspot traffic on a fault-free Mesh $4\times4$, the proposed port-scoring policy outperforms Planar Adaptive routing by 5--16\% and completely eliminates the latency penalties previously observed under transpose traffic. Crucially, the model exhibits architectural resilience: under a severe 15\% random link failure scenario, GNNocRoute sustains a mere 2.5\% latency degradation with zero throughput loss, directly circumventing the catastrophic +257\% latency surge experienced by traditional Planar Adaptive routing. Furthermore, the learned structural representations demonstrate cross-scale transferability, enabling zero-shot transfer from $4\times4$ to $8\times8$ meshes with 99.78\% minimal-port routing accuracy, providing a strong initialization for few-shot fine-tuning that can reduce training epochs by an estimated 60--80\%.

The proposed periodic optimization framework provides a practical, deployable pathway for AI-driven NoC. By executing the GNN encoder asynchronously ($5.5$ ms on CPU) and updating the lightweight weight tables only when macroscopic network conditions shift, GNNocRoute guarantees wire-speed, deadlock-free routing with strictly zero per-packet inference latency overhead.

\subsection*{Future Work}

These results establish a robust foundation for next-generation AI-driven routing, motivating several avenues for future research:

\begin{enumerate}
    \item Transitioning from static precomputed port scores to fully online DRL agents that periodically update routing weights based on real-time microsecond-level congestion observations.
    \item Developing advanced fault-aware training regimes that explicitly model compounded pathological scenarios where severe traffic hotspots intersect with dense link failures.
    \item Validating the framework against trace-driven full-system workloads (e.g., PARSEC, SPLASH-2) within closed-loop execution environments.
    \item Integrating dynamic per-packet energy consumption into the DRL reward function for Pareto-optimal thermal-latency balancing.
    \item Exploring model quantization (FP16/INT8) and dedicated hardware accelerators to further compress the offline GNN inference latency for ultra-high-frequency policy updates.
    \item \textbf{Probabilistic port selection:} Extending the current hardmax port selection to a probabilistic routing scheme where port scores define a probability distribution over minimal directions. This would enable stochastic load balancing under high-utilization scenarios, mitigating the path imbalance observed under uniform traffic at intermediate injection rates.
\end{enumerate}

\subsection*{Acknowledgments}
This research was supported by the Institute of Information Technology, Vietnam National University, Hanoi, under PhD research program No. 25218003.
