% ============================================================
% MASTER FILE — GNNocRoute-DRL Paper
% Journal of Systems Architecture (Elsevier)
% ============================================================
\documentclass[5p,twocolumn]{elsarticle}

\graphicspath{{./figures/}}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{footnote}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage{xcolor}

\journal{Journal of Systems Architecture}

\begin{document}

\begin{frontmatter}

\title{GNNocRoute-DRL: Topology-Aware Adaptive Routing for Network-on-Chip using Graph Neural Networks and Deep Reinforcement Learning}

\author[1]{Le Tien Hieu\corref{cor1}}
\ead{letienhieu@mcs.edu.vn}
\cortext[cor1]{Corresponding author}

\address[1]{Institute of Information Technology, Vietnam National University, Hanoi, Vietnam}

\begin{abstract}
Network-on-Chip (NoC) has become the standard communication infrastructure for modern multi-core processors and AI accelerators. Conventional deterministic routing suffers from structural congestion imbalance, wherein central nodes experience $2-3\times$ higher betweenness centrality than edge nodes. While adaptive routing mitigates congestion under non-uniform traffic, it relies heavily on local heuristics that fail to leverage global topology structure.

This paper presents GNNocRoute-DRL, a framework that decouples offline Graph Neural Network (GNN) inference from online per-packet routing through precomputed port-scoring tables. A GATv2 encoder learns topology-aware node embeddings that correlate strongly with betweenness centrality ($|r| = 0.978$), which are subsequently projected into 4-directional port scores for each source-destination pair. At routing time, each router selects the minimal port with the highest precomputed score via a single-cycle table lookup, introducing a negligible 0.13\% power overhead to the router budget utilizing a 4 KB weight table.

Extensive cycle-accurate evaluations demonstrate that in fault-free $4\times4$ meshes, GNNocRoute reduces hotspot latency by 5--16\% over Planar Adaptive routing and completely eliminates the 115\% transpose traffic penalty observed in prior GNN-based approaches. Crucially, under a severe 15\% random link failure scenario, a fault-aware variant degrades latency by only 2.5\% (compared to a catastrophic +257\% surge for Planar Adaptive) with zero throughput loss. Furthermore, the model exhibits robust zero-shot transferability from $4\times4$ to $8\times8$ meshes with 99.78\% minimal-port accuracy, outperforming XY routing by up to 25\% under hotspot traffic. Compared to state-of-the-art DRL-based NoC routing architectures (MAAR, DeepNR, GARN), GNNocRoute consumes merely 0.3--2\% of their per-hop inference latency while providing superior topology awareness and fault tolerance.
\end{abstract}

\begin{keyword}
Network-on-Chip \sep Graph Neural Networks \sep Port-Score Routing \sep Fault-Tolerant Routing \sep Zero-Shot Generalization
\end{keyword}

\end{frontmatter}

\input{01-introduction}
\input{02-related-work}
\input{03-method}
\input{04-experiments}
\input{05-results}
\input{06-discussion}
\input{07-conclusion}


\nocite{*}
\bibliographystyle{elsarticle-num}
\bibliography{references}

\end{document}
